A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2016-08-13 imaging_demo/ms.tex
Add section on imaging with heterogeneous arrays
From e6e6612 (mnras):
====================

(*) Add section on imaging with heterogeneous arrays
(*) Make associated changes in Abstract, Introduction and Conclusions

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\Nant}{N_\textrm{a}}
\newcommand{\Ngrid}{N_\textrm{g}} 
\newcommand{\Npix}{N_{\text{pix}}}
\newcommand{\dif}{\mathrm{d}}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with 
large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of 
traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep 
cost for realizing the full imaging potential of these powerful instruments. 
Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), 
we present the first software demonstration of a generalized direct imaging 
algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost for dense layouts to 
$\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged 
heterogeneous antenna. EPIC is highly modular and parallelizable, implemented in 
object oriented Python, and publicly available. We have verified the images 
produced to be equivalent to those produced using traditional techniques to 
within a precision determined by coarseness of gridding. We have also validated 
our implementation on data observed with the Long Wavelength Array (LWA). Antenna 
layouts with a dense filling factor consisting of a large number of antennas such 
as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and 
Canadian Hydrogen Intensity Mapping Experiment will gain significant 
computational advantage by deploying EPIC. Inherent availability of calibrated 
time-domain images on digitizer writeout time-scales and vastly lower I/O 
bandwidth relative to visibility-based systems will make it a prime candidate 
for transient searches of Fast Radio Bursts (FRB) as well as planetary and 
exoplanetary phenomena. 
\end{abstract}

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis}, we explore the scalability of the algorithm in the 
context of several array design choices. We identify specific array design 
classes where the EPIC correlator -- is computationally more efficient; and in 
the field of transients, demands significantly lesser I/O bandwidth relative to 
visibility-based approaches. We conclude and discuss future research prospects 
in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. {\bf The noise-like cross-terms ideally have zero mean}. This expression can be further simplified with the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)= \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{W}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power response, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}
V_{ab}(f,t) &= \int \mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by the multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

The intermediate step of gridding with the antenna power response can be expressed as a convolution of a data vector generated by gridding the electric fields directly with the antenna illumination pattern.
\begin{align}
\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v}) &= \left\langle \left[\mathbf{W}_a(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-\mathrm{T}}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right. \nonumber\\ 
&\quad\ast\left.\left[\mathbf{W}^\textrm{T}_a(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

We can then use the multiplication-convolution theorem to move the convolution in Equation \ref{eqn:e-field-conv} to a square after the Fourier transform in Equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-1}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

Equation \ref{eqn:dirty-image-MOFF} is the optimal imaging equation used by the 
MOFF algorithm. While mathematically equivalent to Equation~
\ref{eqn:dirty-image-FX}, squaring in image space rather than convolving in $uv$ 
space potentially saves orders of magnitude in computation.

@@

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility-based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2,-10)$. The reason for these differences is described below.

@@

\section{Imaging with Heterogeneous Antenna Array}\label{sec:versatility}

\begin{figure*}
\subfloat[][Simulated Image]{\label{fig:nonphysical-image}\includegraphics[width=0.5\linewidth]{figure10a.eps}}
\subfloat[][Normalized Radial Flux Density Profile]{\label{fig:norm-flux-density}\includegraphics[width=0.42\linewidth]{figure10b.eps}}
\caption{.}
\label{fig:versatility}
\end{figure*}

>>>>

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\Nant}{N_\textrm{A}}
\newcommand{\Ngrid}{N_\textrm{g}} 
\newcommand{\Npix}{N_{\text{pix}}}
\newcommand{\dif}{\mathrm{d}}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged heterogeneous antenna. We have demonstrated its versatility to image with heterogeneous arrays by showing that it robustly estimates the input sky model while wrong assumptions about array homogeneity lead to severe mis-estimates. EPIC is highly modular and parallelizable, implemented in object oriented Python, and publicly available. We have verified the images produced to be equivalent to those produced using traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying EPIC. Inherent availability of calibrated time-domain images on digitizer writeout time-scales and vastly lower I/O bandwidth relative to visibility-based systems will make it a prime candidate for transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. \S\ref{sec:versatility} demonstrates its ability to robustly image data from heterogeneous arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator -- is computationally more efficient; and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. {\bf The noise-like cross-terms ideally have zero mean}. This expression can be further simplified with the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)= \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{W}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power response, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by the multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

The intermediate step of gridding with the antenna power response can be expressed as a convolution of a data vector generated by gridding the electric fields directly with the antenna illumination pattern.
\begin{align}
\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v}) &= \left\langle \left[\mathbf{W}_a(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}^\textrm{T}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right. \nonumber\\ 
&\,\quad\ast\left.\left[\mathbf{W}^\textrm{T}_a(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}\!(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\right]\right\rangle_t, \label{eqn:e-field-conv}
\end{align}
where, $\mathbf{N}^{-1}$ has been expressed as $\widetilde{\mathbf{N}}^\textrm{T}\widetilde{\mathbf{N}}$.

We can then use the multiplication-convolution theorem to move the convolution in Equation~\ref{eqn:e-field-conv} to a square after the Fourier transform in equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-1}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

Equation~\ref{eqn:dirty-image-MOFF} is the optimal imaging equation used by the MOFF algorithm. While mathematically equivalent to equation~\ref{eqn:dirty-image-FX}, squaring in image space rather than convolving in $uv$ space potentially saves orders of magnitude in computation.

@@

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility-based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2.5,4)$. The reason for these differences is described below.

@@

\section{Imaging with Heterogeneous Antenna Arrays}\label{sec:versatility}

One of the advanced features of the MOFF algorithm and its implementation via EPIC is the ability to naturally account for heterogeneity of antennas while still producing images with maximal information \citep{mor11}. Unlike the assumption frequently made in standard radio interferometry that all antennas are identical, EPIC can implement the most generic case where no such assumption can be made. In other words, EPIC does not assume $W_a(\mathbf{r}-\mathbf{r}_a)$ or $W_{ab}(\mathbf{r}-\mathbf{r}_{ab})$ for different antennas are identical. However, any {\it a priori} knowledge available that certain sets of antennas have identical illumination patterns can be easily passed on to EPIC to avoid redundant computations and thus increase efficiency.

We present a methodology to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array and compare it with that from standard interferometry that assumes a homogeneous array (all antennas are identical). We also demonstrate how the assumption in the latter results in a mis-estimation of recovered flux densities of sources.

Dropping the noise term and time dependence, the visibility measured by an antenna pair can be written from equation~\ref{eqn:measurement-eqn-1} as:
\begin{align}\label{eqn:measurement-eqn-2}
  V_{ab}(f) &= \int \mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\,\mathcal{I}(\hat{\mathbf{s}},f)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, superscript `$\textrm{I}$' in the $\mathcal{W}$ term indicates the weighting is inherently introduced by the instrument during the measurement process. The dirty beam-weighted image is:
\begin{align}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{ab}\, V_{ab}(f)\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}.
\end{align}

When the array is homogeneous, $\mathcal{W}^\textrm{I} \equiv \mathcal{W}^\textrm{I}_a \equiv \mathcal{W}^\textrm{I}_b$ and the above equation reduces to:
\begin{align}\label{eqn:dirty-image-homogeneous}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= |\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f),
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)$ is the dirty image with no beam weighting (isotropic uniform weighting) determined only by the array layout and $|\mathcal{W}^\textrm{I}|^2$ is the directional power pattern of the antenna pair familiar in standard interferometry.

While imaging, EPIC introduces a weighting to each antenna during gridding as given by equations~\ref{eqn:dirty-image-FX} and \ref{eqn:e-field-conv}, and equivalently, the weighted visibility is:
\begin{align}
  V^\prime_{ab}(\mathbf{r},f) &= B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab},
\end{align}
where, $B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)$ is obtained by the spatial cross-correlation of weighting kernels associated with the individual antennas using the definition:
\begin{align}
  B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f) &= \int W^\textrm{G}_a(\mathbf{r}^\prime-\mathbf{r}_a)\,W^{\textrm{G}\star}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_a)\,\dif^2\mathbf{r}^\prime,
\end{align}
and the superscript `$\textrm{G}$' denotes the weighting introduced in analysis during the gridding process. Optimal imaging requires $W^\textrm{G}_a = W^\textrm{I}_a$ \citep{mor09,mor11}. However, we keep the two superscripts to describe output images under the wrong assumption that the array is homogeneous when instead it is actually heterogeneous.

Thus the sky response of the weighted visibility is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \int V^\prime_{ab}(\mathbf{r},f)\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \int B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= V_{ab}\,\int \left[\int W^{\textrm{G}\star}_a(\mathbf{r}^\prime-\mathbf{r}_a)\,W^\textrm{G}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_a)\,\dif^2\mathbf{r}^\prime\right] \nonumber\\
                                              &\qquad\qquad \times e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}
\end{align}
and denotes the fringe from the weighted visibility. The image output of EPIC is equivalent to summing these weighted fringes from all baselines:
\begin{align}\label{eqn:wt-dirty-image-EPIC}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,\mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) \nonumber\\
                                         &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,\left[\mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\right. \nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \times \left. e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\right].
\end{align}

If all antennas are identically weighted ($W^\textrm{G} \equiv W^\textrm{G}_a \equiv W^\textrm{G}_b$), using equation~\ref{eqn:dirty-image-homogeneous}, equation~\ref{eqn:wt-dirty-image-EPIC} reduces to: 
\begin{align}\label{eqn:wt-dirty-image-homogeneous}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{\left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c} \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\left|\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f).
\end{align}
Thus, the dirty image, which is already attenuated by the instrumental power pattern, gets further attenuated by the power pattern introduced in the gridding step in EPIC imager. This is consistent with \citet{mor09}.

We consider a heterogeneous array consisting of a total of $\Nant$ antennas under $N_\textrm{T}$ different types. And under each antenna type $i$, there are $n_i$ antennas such that $\sum_i n_i = \Nant$. The total number of unique non-zero spacing antenna pairs is $\Nant(\Nant-1)/2$ and there are potentially $N_\textrm{T}(N_\textrm{T}+1)/2$ unique types of antenna pairs obtained by pairwise combination of $N_\textrm{T}$ antenna types. Thus,
\begin{align}\label{eqn:total-baselines}
  \frac{1}{2}\Nant(\Nant-1) &= \frac{1}{2}\,\mathop{\sum_{i=1}^{N_\textrm{T}}\sum_{j=1}^{N_\textrm{T}}}_{i\ne j} n_i\,n_j + \frac{1}{2}\,\sum_{i=1}^{N_\textrm{T}} n_i(n_i-1)
\end{align}
is obtained by counting antenna pairs from different antenna types (first term) and those from the same antenna type (second term).

Since equation~\ref{eqn:wt-dirty-image-EPIC} is obtained by summing weighted fringes over all unique antenna pairs, it can be expressed as: 
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\left[\frac{1}{2}\mathop{\sum_{i=1}^{N_\textrm{T}}\sum_{j=1}^{N_\textrm{T}}}_{i\ne j}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}\,\mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f)\right. \nonumber\\ 
  &\qquad\qquad\qquad\qquad\qquad\quad + \left.\frac{1}{2}\sum_{i=1}^{N_\textrm{T}}\sum_{\substack{ab\\a\ne b\\a,b\in i}}\,\mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f)\right]. 
\end{align}
Here, equation~\ref{eqn:wt-dirty-image-EPIC} has been decomposed as the sum of fringes over all unique antenna pairs in an antenna pair type (innermost summation) and subsequently summed over all antenna pair types.

In order to simplify the analysis, we assume that the spatial distribution of antenna pairs in each antenna pair type are similar and thus result in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)$. This usually holds when $\Nant$ is large and antennas under each antenna type are chosen randomly such that differences in the point spread functions are insignificant. Under this assumption, equation~\ref{eqn:wt-dirty-image-EPIC-decomp} can be written as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)}{\Nant(\Nant-1)/2} \nonumber\\
  &\,\times \left[\frac{1}{2}\mathop{\sum_{i=1}^{N_\textrm{T}}\sum_{j=1}^{N_\textrm{T}}}_{i\ne j}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}\left\{\mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\right.\right.\nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \left. \times \mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\right\} \nonumber\\ 
  &\qquad + \frac{1}{2}\sum_{i=1}^{N_\textrm{T}}\sum_{\substack{ab\\a\ne b\\a,b\in i}}\left\{\mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\right. \nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \left.\left.\times\mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\right\}\vphantom{\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}}\right], 
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)$ is approximately the same for all terms and has been pulled outside the summations. For all antennas indexed by $a$ that belong to a particular type $i$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(i)}$ that arise from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. Using arguments similar to those in equation~\ref{eqn:total-baselines}, equation~\ref{eqn:wt-dirty-image-EPIC-decomp-approx1} can be further simplified to:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)}{\Nant(\Nant-1)/2} \nonumber\\
  &\,\times \left[\frac{1}{2}\mathop{\sum_{i=1}^{N_\textrm{T}}\sum_{j=1}^{N_\textrm{T}}}_{i\ne j}n_in_j\left\{\mathcal{W}^{\textrm{G}\star}_{(i)}(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_{(j)}(\hat{\mathbf{s}},f)\right.\right.\nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \left. \times \mathcal{W}^\textrm{I}_{(i)}(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_{(j)}(\hat{\mathbf{s}},f)\right\} \nonumber\\ 
  &\qquad + \frac{1}{2}\sum_{i=1}^{N_\textrm{T}}n_i(n_i-1)\,\left|\mathcal{W}^\textrm{G}_{(i)}(\hat{\mathbf{s}},f)\right|^2\,\left|\mathcal{W}^\textrm{I}_{(i)}(\hat{\mathbf{s}},f)\right|^2\left.\vphantom{\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}}\right].
\end{align}
Then the effective attenuation of the dirty image due to instrumental and gridding weights is:
\begin{align}\label{eqn:effective-weighting}
  \mathcal{W}_\textrm{eff} &= \mathcal{I}^\prime(\hat{\mathbf{s}},f)\, / \,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f).
\end{align}
